WIP: Transparent Standby for Low-Power, Resource-Constrained Embedded Systems

### 2018-02-20 LCTES/body.tex
Correções pontuais da Noemi (e algumas minhas).

According to the International Energy Agency (IEA), the number of
network-connected devices is expected to reach 50 billion by 2020 with the
expansion of the Internet of Things (IoT)~\cite{iea.data}.
%
However, most of the energy to power these devices will be consumed in
\emph{standby mode}, i.e., when they are neither transmitting or processing
data.
%
For instance, standby power currently accounts for 10--15\% of residential
electricity consumption, and $CO_2$ emissions related to standby are equivalent
to those of 1 million cars~\cite{iea.data,standby.australia}.
%
The projected growth of IoT devices, together with the surprising effects of
standby consumption, made network standby efficiency one of the six
pillars of the \emph{G20's Energy Efficiency Action Plan}%
\footnote{G20's Energy Efficiency Action Plan: \url{https://www.iea-4e.org/projects/g20}}.

@@

\CEU is a Esterel-based~\cite{ceu.tecs17} reactive programming language
targeting resource-constrained embedded systems~\cite{ceu.sensys13}.
%
It is grounded on the synchronous concurrency model, which has been
successfully adopted in the context of hard real-time systems such as avionics
and automobiles industry since the 80's~\cite{rp.twelve}.
%
The synchronous model trades power for reliability and has a simpler model
of time that suits most requirements of IoT applications.
%
On the one hand, this model cannot directly express time-consuming
computations, such as compression and cryptography algorithms, which are
typically either absent or delegated to auxiliary chips in the context of the
IoT.
%
On the other hand, all reactions to the external environment are guaranteed to
be computed in bounded time, ensuring that applications always reach an idle
state amenable to standby mode.
%
Overall, \CEU aims to offer a concurrent, safe, and expressive alternative to C
with the characteristics that follow:
%
\begin{description}
\item [Reactive:] code only executes in reactions to events.
\item [Structured:] programs use structured control mechanisms, such as
    \code{await} (to suspend a line of execution), and \code{par} (to combine
    multiple lines of execution).
\item [Synchronous:] reactions run atomically and to completion on each line of
    execution, i.e., there's no implicit preemption or real parallelism.
\end{description}
%
Structured reactive programming lets developers write code in direct style,
recovering from the inversion of control imposed by event-driven
execution~\cite{rp.deprecating,rp.rescala,sync_async.cooperative}.

@@

In order to empower the example in Figure~\ref{lst.inversion}.b with automatic
standby, we developed some extensions to \CEU:
%
\begin{itemize}
\item Make the runtime of \CEU to be interrupt driven and to put the
      microcontroller in standby after each reaction to the environment.
\item Provide operations for the drivers to indicate which interrupts can awake
      the program.
\item Support ISRs in \CEU to generate input events to the program.
\end{itemize}

@@

The driver exposes raw I/O events (ln. 3--4) that only deal with low-level
port manipulation in the microcontroller.
Output events are triggered with the \code{emit} keyword (ln. 29), while input
events are captured with the \code{await} keyword (ln. 30).
%
The output event \code{ADC\_REQUEST} (ln. 9--15) enables ADC interrupts and
starts an analog-to-digital conversion asynchronously in the preipheral for
the single channel \code{A0}.
Any code in between \code{\{} and \code{\}} is treated as an inline $C$ chunk,
allowing for easy integration with $C$ for low-level operations.

@@

The \code{finalize} construct of \CEU executes the nested code whenever its
enclosing block terminates or is aborted externally.
%
The example of Figure~\ref{lst.inversion}.b invokes the \code{AnalogRead}
abstraction (ln. 7) concurrently with \code{RadioAvail} (ln. 2).
%
It is possible that the \code{AnalogRead} terminates normally or that a radio
message arrives during the A/D conversion, which causes the \code{AnalogRead}
to abort abruptly.
In either case, the \code{finalize} clause executes and puts the PMR in a
consistent state.

@@

static u32 pm = 0;

@@

In this work, we address standby efficiency for embedded devices at the level
of programming languages.
%
We propose a software infrastructure for the programming language \CEU that
encompass a power management runtime and support for interrupt service routines
in the language.
%
Our approach relies on the synchronous semantics of the language which enforces
that reactions to the environment always reach an idle state amenable to
standby.
%
We then take energy-unaware programs in \CEU and show that, using the proposed
infrastructure, they can take advantage of the longest periods of time and
deepest sleep modes possible.

>>>>

According to the International Energy Agency (IEA), the number of
network-connected devices is expected to reach 50 billion by 2020 with the
expansion of the Internet of Things (IoT)~\cite{iea.data}.
%
However, most of the energy to power these devices will be consumed in
\emph{standby mode}, i.e., when they are neither transmitting or processing
data.
%
For instance, standby power currently accounts for approximately 10--15\% of
residential electricity consumption, and $CO_2$ emissions related to standby
are equivalent to those of 1 million cars~\cite{iea.data,standby.australia}.
%
The projected growth of IoT devices, together with the surprising effects of
standby consumption, made network standby efficiency one of the six
pillars of the \emph{G20's Energy Efficiency Action Plan}%
\footnote{G20's Energy Efficiency Action Plan: \url{https://www.iea-4e.org/projects/g20}}.

@@

\CEU is a Esterel-based~\cite{ceu.tecs17} reactive programming language
targeting resource-constrained embedded systems~\cite{ceu.sensys13}.
%
It is grounded on the synchronous concurrency model, which has been
successfully adopted in the context of hard real-time systems such as avionics
and automobiles industry since the 80's~\cite{rp.twelve}.
%
The synchronous model trades power for reliability and has a simpler model
of time that suits most requirements of IoT applications.
%
On the one hand, this model cannot directly express time-consuming
computations, such as compression and cryptography algorithms, which are
typically either absent or delegated to auxiliary chips in the context of the
IoT.
%
On the other hand, all reactions to the external environment are guaranteed to
be computed in bounded time, ensuring that applications always reach an idle
state amenable to standby mode.
%
Overall, \CEU aims to offer a concurrent, safe, and expressive alternative to C
with the characteristics that follow:
%
\begin{description}
\item [Reactive:] code only executes in reactions to events and is idle most of
    the time.
\item [Structured:] programs use structured control mechanisms, such as
    \code{await} (to suspend a line of execution), and \code{par} (to combine
    multiple lines of execution).
\item [Synchronous:] reactions run atomically and to completion on each line of
    execution, i.e., there's no implicit preemption or real parallelism.
\end{description}
%
Structured reactive programming lets developers write code in direct style,
recovering from the inversion of control imposed by event-driven
execution~\cite{rp.deprecating,rp.rescala,sync_async.cooperative}.

@@

In order to empower the example in Figure~\ref{lst.inversion}.b with automatic
standby, we developed some extensions to \CEU:
%
\begin{itemize}
\item We made the runtime of \CEU interrupt driven and put the microcontroller
      in standby after each reaction to the environment.
\item We provided operations for the drivers to indicate which interrupts can
      awake the program.
\item We included support for ISRs in \CEU to generate input events to the
      program.
\end{itemize}

@@

The driver exposes raw I/O events (ln. 3--4) that only deal with low-level
port manipulation in the microcontroller.
Output events are triggered with the \code{emit} keyword (ln. 29), while input
events are captured with the \code{await} keyword (ln. 30).
%
The output event \code{ADC\_REQUEST} (ln. 9--15) enables ADC interrupts and
starts an analog-to-digital conversion asynchronously in the peripheral for
the single channel \code{A0}.
In \CEU, any code in between \code{\{} and \code{\}} is treated as an inline
$C$ chunk, allowing for easy integration with $C$ for low-level operations.

@@

The \code{finalize} construct of \CEU executes the nested code whenever its
enclosing block terminates or is aborted externally.
%
The example of Figure~\ref{lst.inversion}.b invokes the \code{AnalogRead}
abstraction (ln. 7) concurrently with \code{RadioAvail} (ln. 2).
%
The \code{AnalogRead} may terminate normally or a radio message may arrive
during the A/D conversion, causing the \code{AnalogRead} to abort abruptly.
In either case, the \code{finalize} clause executes and puts the PMR in a
consistent state.

@@

static u32 pm = 0; // up to 32 peripherals

@@

In this work, we address standby efficiency for embedded devices at the level
of programming languages.
%
We propose a software infrastructure for the programming language \CEU that
encompasses a power management runtime and support for interrupt service
routines in the language.
%
Our approach relies on the synchronous semantics of the language which enforces
that reactions to the environment always reach an idle state amenable to
standby.
%
This way, application written in \CEU can take advantage of the longest periods
of time and deepest sleep modes possible without extra programming efforts.
